---
title: "Lupus"
author: "Reuben"
date: "2025-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Clean environment

```{r}
rm(list = ls())
```

```{r}
library(Seurat)
library(ggplot2)
library(pheatmap)
library(dplyr)
```

Load Seurat object
```{r}
object = readRDS("SkinAnnotatedSeuratObject.RDS")
print(object)
```
Clear the object, as we are going to perform the analysis from scratch, so we only want the raw data
```{r}
# Extract raw counts from the RNA assay
raw_counts <- GetAssayData(object, assay = "RNA", slot = "counts")
# Create a new Seurat object with only raw counts - this will automatically remove counts that have 0 values for all cells
new_object <- CreateSeuratObject(counts = raw_counts)

# copy over the metadata - only the important aspects eg. sample type and percentage of mitochondrial content- we dont want to copy the clustering information because we are doing that ourselves.
new_object@meta.data <- object@meta.data[1:8]

# Check new object
print(new_object) 
```

```{r}
rm(object)
```


 Get an idea of what the raw counts and meta data look like.
```{r}
print(head(new_object@assays$RNA$counts))
print(head(new_object@meta.data))
```

Quality control - we want to remove cells with high mitochondrial gene expression, as this is a sign that the cell is stressed or dying.
We also want to remove cells with a very low or very high amounts of detected genes, as these can represent empty droplets or droplets with more than one cell in the data generation process.
```{r}
#We already have the percentage of mitochondrial gene counts in our metadata.
#fliter out bad cells
new_object <- subset(new_object, subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mito < 10)
```


Normalise between cells to account for differences in library size and read depth.
```{r}
new_object <- NormalizeData(new_object)
```
We can now look for the most varying genes in our dataset, as these will hold the most biological meaning.
```{r}
# Find the genes that vary the most between cells
new_object <- FindVariableFeatures(new_object, selection.method = "vst", nfeatures = 2000)
print(VariableFeaturePlot(new_object))
```


Scale across genes to make them comparable.
```{r}
new_object <- ScaleData(new_object, vars.to.regress = "percent.mito", features = VariableFeatures(new_object))
# regress out the effect of mitochondrial genes
```

Perform a PCA on the processed data
PCA is performed to lower the dimensionality of the dataset.
```{r}
new_object <- RunPCA(new_object, feature = VariableFeatures(new_object))
print(new_object)
```
 
Plot an Elbow plot to identify how many dimensions to include.
```{r}
print(ElbowPlot(new_object))
```
It seems like after 15 dimensions very little variability is captured,
so we will continue using 15 dimensions.

Next we will cluster the data using nearest neighbours.

```{r}
new_object <- FindNeighbors(new_object, dims = 1:15, )
new_object <-FindClusters(new_object, resolution = 0.5)

# plot the clusters in a UMAP
new_object <- RunUMAP(new_object, dims = 1:15)
print(DimPlot(new_object, reduction = 'umap'))
```
```{bash}
#save processed object
saveRDS(new_object, file = "filtered_scaled_seurat_object.rds")
```

```{r}
# plot again labelling based on sample type to determine if we should use harmony integration.
print(DimPlot(new_object, reduction = 'umap', group.by = "SampleGroup"))
print(DimPlot(new_object, reduction = 'umap', group.by = "SampleID"))
```

Based on these two plots it does not appear that harmony is needed as the different samples seem to be integrated, so there isnt a batch effect.


```{r}
#Find marker genes for each cluster
new_object.markers <- FindAllMarkers(new_object, 
                                only.pos = TRUE,
                                min.pct = 0.25,
                                logfc.threshold = 0.5)
#Extract top 3 markers for each cluster based on avg log2 fold change
top_markers <- new_object.markers %>% 
    group_by(cluster) %>%
    slice_max(n = 5, order_by = avg_log2FC)
print(top_markers)

#Plot a single top marker gene for the first cluster using the UMAP too see what cells have a high expression level,
print(FeaturePlot(new_object, features = top_markers$gene[1]))
#We can see that high expression is mainly in the cluster 0 for this gene, which is expected. Now we need to determine what cell types each marker gene is associated with.
```
```{r}
cell_markers <- list(
  Activated_T_cell = c("CD69", "IL2", "IFNG"),
  Adipocyte = c("APOE", "CFD", "ADIPOQ"),
  Antigen_presenting_cell = c("HLA-DRA", "CD74", "HLA-DPB1"),
  B_cell = c("CD19", "MS4A1", "CD79A"),
  Basal_cell = c("KRT14", "KRT5", "CDH3"),
  CD4_T_cell = c("CD3E", "CD4", "IL7R"),
  CD8_T_cell = c("CD8A", "CD8B", "GZMK"),
  Central_memory_T_cell = c("CCR7", "SELL", "CD45RO"),
  Conventional_dendritic_cell_1_cDC1 = c("IRF8", "CLEC9A", "BATF3"),
  Conventional_dendritic_cell_2_cDC2 = c("CD1C", "IRF4", "FCER1A"),
  Dendritic_cell = c("ITGAX", "CD1C", "HLA-DRA"),
  Dermal_cell = c("APCDD1", "SFRP2", "WIF1"),
  Differentiated_keratinocyte = c("KRT1", "KRT10", "IVL"),
  Dopamine_neuron = c("TH", "DDC", "SLC18A2"),
  Endothelial_cell = c("PECAM1", "VWF", "CDH5"),
  Epidermal_cell = c("EPCAM", "IVL", "TP63"),
  Fibroblast = c("COL1A1", "FAP", "PDGFRA"),
  Gamma_delta_T_cell = c("TRDC", "TRGC1", "NCAM1"),
  Healing_Enriched_Fibroblast = c("PLA2G2A", "COL1A1", "FBLN1"),
  Keratinocyte = c("KRT14", "KRT5", "SPRR2A"),
  Lymphatic_endothelial_cell = c("LYVE1", "PROX1", "PDPN"),
  Lymphocyte = c("CD3D", "CD7", "CD2"),
  M1_macrophage = c("IL1B", "S100A8", "TNF"),
  M2_macrophage = c("CD163", "MRC1", "CD204"),
  Macrophage = c("CD68", "CD163", "LYZ"),
  Mast_cell = c("TPSAB1", "TPSB2", "CPA3"),
  Melanocyte = c("MLANA", "PMEL", "TYR"),
  Memory_B_cell = c("MS4A1", "CD27", "CD19"),
  Memory_CD4plus_T_cell = c("CD4", "IL7R", "SELL"),
  Memory_T_cell = c("CD28", "SELL", "IL7R"),
  Monocyte = c("CD14", "LYZ", "FCN1"),
  Myeloid_cell = c("LYZ", "CD68", "CXCL8"),
  Myeloid_dendritic_cell = c("CD86", "CD40", "CD209"),
  Natural_killer_cell = c("NCAM1", "NKG7", "GZMB"),
  Natural_killer_T_cell = c("CD3E", "NKG7", "CCL5"),
  Papillary_fibroblast = c("PDPN", "COL6A1", "APCDD1"),
  Regulatory_T_cell = c("CD3E", "CTLA4", "FOXP3"),
  Resident_memory_T_cell = c("CD69", "ITGAE", "CXCR6"),
  Schwann_cell = c("SOX10", "PLP1", "S100B"),
  Smooth_muscle_cell = c("ACTA2", "TAGLN", "MYH11"),
  Sweat_and_sebaceous_gland_cell = c("DCD", "AQP5", "KRT7"),
  T_cell = c("CD3D", "CD3E", "CD2"),
  Tissue_resident_memory_T_cell = c("CD69", "ITGAE", "CXCR6")
)

cluster_cell_types <- list()

#Loop through each cluster number
for (cluster_num in unique(top_markers$cluster)) {
  
  #Get the top marker genes for this cluster
  top_genes <- top_markers[top_markers$cluster == cluster_num, ]$gene
  
  #Initialize an empty vector to hold matching cell types for this cluster
  cell_type_matches <- c()
  
  #Loop through each cell type in the marker list
  for (cell_type in names(cell_markers)) {
    genes <- cell_markers[[cell_type]]
    
    #Check if any of the top marker genes match the known markers for this cell type
    if (length(intersect(top_genes, genes)) > 0) {
      #Add the cell type name to the matches for this cluster
      cell_type_matches <- c(cell_type_matches, cell_type)
    }
  }
  
  #Add the cell types for this cluster into the main list
  cluster_cell_types[[as.character(cluster_num)]] <- cell_type_matches
}

# Now print or inspect the matched cell types for each cluster
print(cluster_cell_types)
```
```{r}
#plot clusters for comparison
print(DimPlot(new_object, reduction = 'umap'))
```

#Now we have potential cell labels for each cluster. To finalise these we should plot them to check.
```{r}
#visualise marker expression for the two candidate cell types in cluster 0.
print(FeaturePlot(new_object, features = cell_markers[["Regulatory_T_cell"]]))
print(FeaturePlot(new_object, features = cell_markers[["Memory_T_cell"]]))
#Cluster 0 shows stronger expression of Regulatory T cell markers, 
#while Memory T cell markers are distributed across other clusters.
#We'll label cluster 0 as Regulatory T cells.
```

```{r}
#visualise marker expression for the two candidate cell types in cluster 1.
print(FeaturePlot(new_object, features = cell_markers[["Healing_Enriched_Fibroblast"]]))
#A lot of these markers also overlap with cluster 4, so I am not confident to label this cluster  #Healing_Enriched_Fibroblast, however, it is likely some type of fiberblast still, so labelling 
#it fiberblast will be ok for now.
```

```{r}
#cluster 2 was not found with our marker list, so we will move to cluster 3.
print(FeaturePlot(new_object, features = cell_markers[[ "Smooth_muscle_cell"]]))
# cluster 3 definitely seems to be a smooth muscle cell.
```
```{r}
#cluster 4.
print(FeaturePlot(new_object, features = cell_markers[["Dermal_cell"]]))
# cluster 4 definitely seems to be a dermal cell.
```

```{r}
#clusteer 5 was also missed out so we will look at cluster 6.

print(FeaturePlot(new_object, features = cell_markers[["M1_macrophage"]]))

print(FeaturePlot(new_object, features = cell_markers[["Macrophage"]]))

print(FeaturePlot(new_object, features = cell_markers[["Monocyte"]]))

print(FeaturePlot(new_object, features = cell_markers[["Myeloid_cell"]]))
# likely a Macrophage or myeloid cel
# This is not supprising as macrophages are a specialised type of myeloid cells.
```
```{r}
# cluster 7 definitely a CD8 T cell
print(FeaturePlot(new_object, features = cell_markers[["CD8_T_cell"]]))
```
```{r}
# cluster 8 definitely a Differentiated keratinocyte
print(FeaturePlot(new_object, features = cell_markers[["Differentiated_keratinocyte"]]))

```
```{r}
# cluster 9 missed, cluster 10 definitely a Melanocyte
print(FeaturePlot(new_object, features = cell_markers[["Melanocyte"]]))

```
```{r}
# cluster 11 definitely a Lymphatic endothelial cell
print(FeaturePlot(new_object, features = cell_markers[["Lymphatic_endothelial_cell"]]))
```
```{r}

# cluster 12 definitely Sweat and sebaceous gland cells
print(FeaturePlot(new_object, features = cell_markers[["Sweat_and_sebaceous_gland_cell"]]))
```
```{r}
#clusters 13 and 15 are missed
# cluster 14, definitely a B cell.
print(FeaturePlot(new_object, features = cell_markers[["B_cell"]]))

print(FeaturePlot(new_object, features = cell_markers[["Memory_B_cell"]]))
```

